Adapting the Chumbley Score to Bullet Striations

### 2018-04-24 presentation/presentation.Rmd
presentation modifications

## Data
-  Ruger P85s Bullet Lands, or Hamby scans (\cite{hamby}) provided by NIST  (85,491 comparisons)
-  Bullet striation marks $\approx$ 2mm | Screwdriver marks $\approx$ 7mm (all chumbley score papers)

### Getting from Profiles to Signatures
\vspace{-0.5em}

- Lowess smoothing or Loess fit residuals $=$ Signatures
- Removes topographic structure (curvature)
- Improve the signal to noise Ratio

@@

## Error Rates for toolmarks

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{\citet{manytoolmarks1}} & Maximized & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{\citet{chumbley}} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{\citet{afte-chumbley}} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{\citet{hadler}} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
 \textbf{\citet{manytoolmarks2}} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates for Toolmarks}
\end{table}

## Digitized Striation Marks

- Signatures/ Profiles (NIST- Hamby) $\approx$ 1200 pixels (2 mm)
  Screwdriver toolmarks (Chumbley Papers) $\approx$ 9000 pixels (7 mm)
- Markings are smoothed as a preprocessing step
    - Used to remove drift and (sub)class characteristics from individual markings

## Overview

\caption{\label{flow-chart} An overview of the chumbley score method}
\end{figure}

- Two steps: Optimization ($1^{st}$) and Validation ($2^{nd}$).
- Windows $\implies$ short segments of the markings
    - Have \textit{predefined sizes}. ($T_1$ or $T_2 >>> w_o >> w_v$)
          1. $w_o$ used in the Optimization step 
          2. $w_v$ used in the Validation step

@@

- Aligns markings horizontally 
- Correlation of all possible windows of size $w_o$ between $x(t_1)$ and $y(t_2)$ computed
- \textit{Possible lag in the 2 markings} \newline Window Pair with maximized correlation $\implies$ \newline Optimal vertical (in-phase) shift of $t_1^o - t_2^o$ 
      - For aligning the two markings. 
\vspace{-1em}
\[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1, 1 \le t_2 \le T_2} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\]

@@

#### Validations Step
- Two sets of windows of size $w_v$ chosen from both markings (see Figure \ref{fig:win-comparison})

@@

## 2

More precisely, let us define starting points of the windows of validation $s_i^{(k)}$ for each signature $k = 1, 2$ as 
\begin{eqnarray}\label{eq.start}
s^{(k)}_i = 
\begin{cases}
t_k^* + i w_v^* & \text{ for } i < 0 \\
t_k^* + w_ o^* + i w_v^* & \text{ for } i \ge 0,
\end{cases}
\end{eqnarray} for integer values of $i$ with $0 <  s^{(k)}_i \le 1 - w_v^*$
<!-- $0 <  s^{(k)}_i \le T_k - w_v$. -->

## 3

- Same-shift pairs of length $w_v$ are all pairs $$(s_i^{(1)}, s_i^{(2)}) \quad \forall \, i \, \in \, \mathbb{Z}$$ for which both $s_i^{(1)}$ and  $s_i^{(2)}$ are defined. 
- Different-shift pairs are defined as $$(s_i^{(1)}, s_{-i-1}^{(2)}) \quad \forall \, i \, \in \, \mathbb{Z}$$ where both $s_i^{(1)}$ and  $s_{-i-1}^{(2)}$ are defined (see \cref{sketch-same-diff}).

@@

## 4
- No overlaps of pairs within selected marks $\implies$ independence.
- Both same- and different-shift pairs correlations between the  markings are calculated. 
- For Same-Source markings correlations
    - for the in-phase shift should be high
    - for out-of-phase shift should be low.
          - Provide a measure for the base-level correlation to which in-phase shift correlations can be compared.
- The Chumbley score is the Mann Whitney U statistic computed by comparing between in-phase sample and out-of-phase sample.

@@

- By definition (equation \autoref{eq.start}), some number of tests fail to produce a result
- Either because the number of eligible same-shift pairs is 0, or the number of different-shift pairs is 0.

\vspace{-0.5em}

###### No Same-shift pairs if
- the optimal locations $t_1^{o}$ and $t_2^{o}$ are in the extremes
    - no segments of size $w_v$ are left on the same sides of the optimal locations, 
    - For all $t_1^{o} < w_v$ and $t_2^{o} > T_2-w_o-w_v$ or \newline $t_1^{o} < T_1- w_o - w_v$ and $t_2^{o} < w_v$

\vspace{-0.5em}

###### No Different Shift Pairs if
- the optimal locations are close to the boundaries
    - \# of different-shift pairs also depends on the placement of the  optimal locations $t_i^o$.
    -  For all $t_i^{(o)} < w_v$ or  $t_i^{(o)} > T_i-w_o- w_v$.

\vspace{-0.5em}

@@

#### Same-shift failure

@@

## 6

- For different-source matches
    - we can assume $t_1^o$ and $t_2^o$ are independent.
    - Failure rate = $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$ 
    - For an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels this probability is about 0.0015.
- For same-source matches, 
    - we expect a strong dependency between optimal locations $t_i^o$
    - $t_1^o$ and $t_2^o$ are not independent of each other
    - A large difference between locations is unlikely
    - Failure Rate = $2 w_v / (T_i-w_o)$, or about 5.6\% for an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels.

##  Failure rate for same-source matches
The failure rate for same-source matches can also be written as follows
$$
\begin{split}
P\left( t_1^{o} < w_v \ \bigcup \ t_2^{o} > T_2-w_o-w_v\right) + \\P\left( t_1^{o} < T_1- w_o - w_v \ \bigcup \ t_2^{o} < w_v\right)
\end{split}
$$

Taking first part of the above equation, let $A = t_1^{o} < w_v$ and $B = t_2^{o} > T_2-w_o-w_v$

- Again we assume high correlation between optimal locations
- In order to have at least one same-shift pair or different shift pair
    - \textbf{Either}, if $x \in A$, then $B$ is empty and $P(A \cup B) = P (A)$
    - \textbf{or}, if $x \in B$, then $A$ is empty and $P(A \cup B) = P(B)$

This would hold for the second part too, and would therefore give an expected rate of failure as $2 w_v / (T_i-w_o)$

>>>>

## Variations of Chumbley score method and Error Rates for toolmarks

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{\citet{manytoolmarks1}} & Maximized & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{\citet{chumbley}} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{\citet{afte-chumbley}} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{\citet{hadler}} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ 
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates for Toolmarks using  variations of the chumbley score method}
\end{table}

 <!-- \textbf{\citet{manytoolmarks2}} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\ -->

## Digitized Striation Marks
- Data
    -  Ruger P85s Bullet Lands, or Hamby scans (\cite{hamby}) provided by NIST  (85,491 comparisons)
    -  Bullet striation marks $\approx$ 2mm 
    -  Screwdriver marks $\approx$ 7mm (all chumbley score papers)

- Signatures/ Profiles (NIST- Hamby) $\approx$ 1200 pixels (2 mm)
  Screwdriver toolmarks (Chumbley Papers) $\approx$ 9000 pixels (7 mm)

#### Step 0 : Defining a coarseness parameter
- Used to remove drift and (sub)class characteristics from individual markings
<!-- ### Getting from Profiles to Signatures -->
<!-- \vspace{-0.5em} -->
- Lowess or Loess fit residuals $=$ Signatures
- Removes topographic structure (curvature)
- Improve the signal to noise Ratio

@@

## Algorithm

- Two steps: Optimization ($1^{st}$) and Validation ($2^{nd}$).
- Windows $\implies$ short segments of the markings
    - Have \textit{predefined sizes}. ($T_1$ or $T_2$ >>> $w_o$ & w_v$)
          1. $w_o$ used in the Optimization step 
          2. $w_v$ used in the Validation step

@@

- \textbf{Goal :}Align markings horizontally as best as possible
- Correlation Matrix of all possible windows of size $w_o$ between $x(t_1)$ and $y(t_2)$ computed
- \textit{Identify lag for horizontal alignment} \newline Window Pair with maximized correlation $\implies$ \newline Optimal vertical (in-phase) shift of $t_1^o - t_2^o$ 
      - For aligning the two markings. 
\vspace{-1em}
\[
\left(t_1^o, t_2^o\right) = \mathop{\arg \max}\limits_{1 \le t_1 \le T_1, 1 \le t_2 \le T_2} \text{cor} \left(x^{w_o} (t_1), y^{w_o}(t_2) \right)
\]

@@

#### Validation Step
- Two sets of windows of size $w_v$ chosen from both markings (see Figure \ref{fig:win-comparison})

@@

## 4
<!-- - No overlaps of pairs within selected marks $\implies$ independence. -->
- Both same- and different-shift pairs correlations between the  markings are calculated. 
- For Same-Source markings, correlations
    - for the in-phase shift should be high
    - for out-of-phase shift should be low.
          - Provide a measure for the base-level correlation to which in-phase shift correlations can be compared.
- The Chumbley score is the Mann Whitney U statistic computed by comparing between in-phase sample and out-of-phase sample.

## Block Diagram

\caption{\label{flow-chart} An overview of the adjusted chumbley score method as given by \citet{hadler}}
\end{figure}

## Starting Points

More precisely, let us define starting points of the windows of validation $s_i^{(k)}$ for each marking $k = 1, 2$ as 
\begin{eqnarray}\label{eq.start}
s^{(k)}_i = 
\begin{cases}
t_k^* + i w_v^* & \text{ for } i < 0 \\
t_k^* + w_ o^* + i w_v^* & \text{ for } i \ge 0,
\end{cases}
\end{eqnarray} for integer values of $i$ with $0 <  s^{(k)}_i \le 1 - w_v^*$
<!-- $0 <  s^{(k)}_i \le T_k - w_v$. -->

## The \citet{hadler} method (CS1)

- Same-shift pairs of length $w_v$ are all pairs that start in: $$(s_i^{(1)}, s_i^{(2)}) \quad \forall \, i \, \in \, \mathbb{Z}$$ for which both $s_i^{(1)}$ and  $s_i^{(2)}$ are defined. 
- Different-shift pairs are defined as $$(s_i^{(1)}, s_{-i-1}^{(2)}) \quad \forall \, i \, \in \, \mathbb{Z}$$ where both $s_i^{(1)}$ and  $s_{-i-1}^{(2)}$ are defined (see \cref{sketch-same-diff}).

@@

@@

- By definition (equation \autoref{eq.start}), some number of tests fail to produce a result
- Either because the number of eligible same-shift pairs is 0, or the number of different-shift pairs is 0. 
- $t_1^o, t_2^o$ not necessarily independent
    - \textbf{same-source:} Assume high dependence, corr($t_1^o,t_2^o$) $\approx$ 1
        - Example: $w_o$ = 120, coarseness (c) = 0.3, corr($t_1^o,t_2^o$) = 0.85
    - \textbf{diff-source:} Assume independence of $t_1^o,t_2^o$
        - Example: $w_o$ = 120, coarseness (c) = 0.3, corr($t_1^o,t_2^o$) = 0.12
<!-- \vspace{-0.5em} -->

<!-- ###### No Same-shift pairs if -->
<!-- - the optimal locations $t_1^{o}$ and $t_2^{o}$ are in the extremes -->
<!--     - no segments of size $w_v$ are left on the same sides of the optimal locations,  -->
<!--     - For all $t_1^{o} < w_v$ and $t_2^{o} > T_2-w_o-w_v$ or \newline $t_1^{o} < T_1- w_o - w_v$ and $t_2^{o} < w_v$ -->

<!-- \vspace{-0.5em} -->

<!-- ###### No Different Shift Pairs if -->
<!-- - the optimal locations are close to the boundaries -->
<!--     - \# of different-shift pairs also depends on the placement of the  optimal locations $t_i^o$. -->
<!--     -  For all $t_i^{(o)} < w_v$ or  $t_i^{(o)} > T_i-w_o- w_v$. -->

<!-- \vspace{-0.5em} -->

@@

## Same-shift failure
- Same-source $\approx$ 0
- Different-source $\approx$ 2 $P(t_i < w_o)^2$ = $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$

@@

## Different-Shift Failure

- Same-source (Assuming $t_1^{o} \approx t_2^{o}$) $\approx$ $2w_v / (T_i-w_o)$\newline

\vspace{-2em}

$P( t_1^{o} < w_v \ \cap \ t_2^{o} < w_v) + P( t_1^{o} <  w_v \ \cap \ t_2^{o} < w_v)  = \ 2P(t_o^1 \ < \ w_v)$

- Different-source $\approx$  $2P(t_i < w_o)^2$ = $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$

\begin{figure}[hbtp]
\centering
\includegraphics[width=0.8\textwidth]{imgs/diff-shift-failure.png}

\caption{\label{diff-shift-failure}Sketch of diff-shift pairings  (top) when the number of diff-shift computations is likely to 0}
\end{figure}

<!-- - For different-source matches -->
<!--     - we can assume $t_1^o$ and $t_2^o$ are independent. -->
<!--     - Failure rate = $\frac{2 w_v^2}{(T_1-w_o)(T_2-w_o)}$  -->
<!--     - For an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels this probability is about 0.0015. -->
<!-- - For same-source matches,  -->
<!--     - we expect a strong dependency between optimal locations $t_i^o$ -->
<!--     - $t_1^o$ and $t_2^o$ are not independent of each other -->
<!--     - A large difference between locations is unlikely -->
<!--     - Failure Rate = $2 w_v / (T_i-w_o)$, or about 5.6\% for an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels. -->

<!-- ##  Failure rate for same-source matches -->
<!-- The failure rate for same-source matches can also be written as follows -->
<!-- $$ -->
<!-- \begin{split} -->
<!-- P\left( t_1^{o} < w_v \ \bigcup \ t_2^{o} > T_2-w_o-w_v\right) + \\P\left( t_1^{o} < T_1- w_o - w_v \ \bigcup \ t_2^{o} < w_v\right) -->
<!-- \end{split} -->
<!-- $$ -->

<!-- Taking first part of the above equation, let $A = t_1^{o} < w_v$ and $B = t_2^{o} > T_2-w_o-w_v$ -->

<!-- - Again we assume high correlation between optimal locations -->
<!-- - In order to have at least one same-shift pair or different shift pair -->
<!--     - \textbf{Either}, if $x \in A$, then $B$ is empty and $P(A \cup B) = P (A)$ -->
<!--     - \textbf{or}, if $x \in B$, then $A$ is empty and $P(A \cup B) = P(B)$ -->

<!-- This would hold for the second part too, and would therefore give an expected rate of failure as $2 w_v / (T_i-w_o)$ -->
